Interpreting Random Forest Predictions for Firearm Identification Using LIME

### 2018-10-23 paper/paper.Rmd
worked more on introduction

The discipline of firearm identification examines bullets to determine the likelihood that a bullet found in a criminal case was fired from a particular gun. This is a procedure that has traditionally been performed by hand. Specially trained examiners visually assess the striations on the unknown bullet from the crime that were created when the bullet passed through the gun barrel to those on a known bullet that was fired from the gun under evalution. Often a comparison microscope is used that allows the examiners to view both bullets at the same time (National Research Counci 2009). The examiners use the ability to compare the two bullets to determine whether the two bullets were fired from the same gun barrel.

Recently, the scientific community has been encouraging the inclusion of more data driven techniques to be used in forensic investigations that would allow for the reporting of a measure of uncertainty in addition to the conclusion drawn from the analysis. This led Hare, Hofmann, and Carriquiry (2017) to propose a new computer automated method of bullet matching. The method involves obtaining bullet signatures from the scans of the two bullets, computing a handful of variables that measure the similarity between two signatures, and fitting a random forest model to the similarity variables. The model can be trained on a set of known bullet matches and non-matches. Then it can be used to predict on a new set of bullet signature comparisons. The results from the paper suggest that the random forest model leads to highly accurate bullet matching predictions. The authors found that when their model was applied to a testing dataset (?), the resulting error rate was 0%.

While random forest models often result in good predictions, it is well known that a disadvantage of random forest models and other machine learning techniques is that it is difficult to interpret the models (reference?). For example, it is not possible to tell which variables played an important role in the creation of individual predictions. This issue led to the development of LIME (reference), which is an algorithm that examines the behavior of the complicated model on a local scale around a new prediction using a linear model. This allows for the ability to understand which were the driving variables in a prediction of interest.

Since firearm identification identification is commonly used as evidence for convictions in court cases, it is important to not only provide accurate predictions and an associated uncertainty level but also to understand which variables were important in making the prediction. Firearm examiners could use LIME to check whether or not the predictions created by LIME are reasonable. This paper provides an example of the application of LIME to an updated version of the random forest model from the Hare paper used for bullet matching.

>>>>

(need to finish adding sources)

The discipline of firearm identification examines bullets to determine the likelihood that a bullet found in a criminal case was fired from a particular gun. To do this, the bullet from the crime will be compared with a bullet that was known to be fired from the gun under evaluation. Traditionally, this is a procedure that has been performed by hand. Specially trained examiners visually compare the microscopoic bullet striations that were created when the bullets passed through the gun barrel. Often a comparison microscope is used that allows the examiners to view both bullets at the same time (National Research Council 2009). The examiners use this ability to determine whether the two bullets were fired from the same gun barrel.

Recently, the scientific community has been encouraging the inclusion of more data driven techniques to be used in forensic investigations. These methods would allow for the reporting of a measure of uncertainty in addition to the conclusion drawn from the analysis. This led Hare, Hofmann, and Carriquiry (2017) to propose a new computer automated method of bullet matching that could supplement the visual inspection by the firearm examiners. Their method involves obtaining bullet signatures of the striations from the scans of the two bullets, computing variables that measure the similarity of the two signatures, and using a trained random forest model to determine the probability of a match based on the similarity variables. They trained their model on a set of known bullet matches and non-matches from the Hamby study. They demonstrated how the random forest model can be used to make prediction on a new set of bullet signature comparisons. The results from the paper suggest that the random forest model leads to highly accurate bullet matching predictions. The authors found that when their model was applied to a testing dataset (?), the resulting error rate was 0%.

While random forest models often result in good predictions as seen in Hare, Hofmann, and Carriquiry (2017), it is well known that a disadvantage of random forest models and other machine learning techniques is that it is difficult to interpret the models (reference?). For example, it is not possible to tell which variables played an important role in the creation of individual predictions. This issue led to the development of LIME (reference), which is an algorithm that examines the behavior of the complicated model on a local scale around a new prediction using a linear regression model. This allows for the ability to understand which were the driving variables that led to a prediction of interest.

Since firearm identification identification is commonly used as evidence for convictions in court cases, it is important to be able to understand and assess the model that is being used to quantify the probability that a bullet was fired from a gun. LIME provides the ability to understand which were the key variables used by the random forest model to make a prediction, which would allow firearm examiners to check whether or not the predictions created by the random forest are based on reasonable variables.

This paper provides an example of the application of LIME to a bullet matching problem. (provide more details about what is contained in the papers - i.e. section 2 describes the Hamby data; section 3 describes the random forest model and LIME; ...)
